A Standardized Effect Size for Evaluating and Comparing the Strength of Phylogenetic Signal

### 2020-08-12 Manuscript/PNAS-Adams-02.tex
vsn 2

The appeal of Pagel's \(\lambda\) and Blomberg's \(\kappa\) as
descriptive statistics is that they provide a basis for interpreting
``weak'' versus ``strong'' phylogenetic signal; i.e., small versus large
values of \(\hat{\lambda}\) or \(\kappa\), respectively, in a
comparative sense (28--30). Nonetheless, an important question that has
yet to be considered is whether these statistics are or can be converted
to effect sizes for comparative analyses across datasets? To be
statistics representing phylogenetic signal, they should have reliable
distributional properties, which could be revealed with simulation
experiments. For instance, as a proportional random variable bounded by
\(0\) and \(1\), we might expect that \(\hat{\lambda}\) follows a
Bernoulli distribution (\textbf{add ref}); i.e., branch lengths in a
tree are scaled proportionally to the probability that data arise from a
BM process. Given a known \(\lambda\) value used to generate random data
on a tree, we would also expect that the mean of an empirical sampling
distribution of \(\hat{\lambda}\) would approximately equal \(\lambda\);
the dispersion of \(\hat{\lambda}\) would be largest at intermediate
values of \(\lambda\), \(\hat{\lambda}\) would be predictable over the
range of \(\lambda\) with respect to tree size; the distribution of
\(\hat{\lambda}\) would be symmetric at intermediate values of
\(\lambda\) and more skewed toward values of 0 or 1; and that the
distribution of \(\hat{\lambda}\) will be more platykurtic at
intermediate values of \(\lambda\), becoming more leptokurtic toward 0
and 1 (\textbf{add same ref}). Prior work (18) seems to support some of
these conjectures, based superficially on statistical moments for a
given tree size (mean, variance, skewness, and kurtosis; see Fig. 2 of
ref. (18)). However, because the ``strength of Brownian motion'' was
simulated as a varied weighted-average of data simulated on trees with
\(\lambda=0\) and \(\lambda=1\) and not as prescribed values of
\(\lambda\) (18), interpretation of these patterns is challenging.

By contrast, for Blomberg's \(\kappa\), which is positively uNbounded,
we might expect that for any \(\lambda\) used to generate data,
estimates of \(\kappa\) might follow a normal distribution, with values
distributed symmetrically about the input value. This attribute seemed
less reasonable based on the simulations performed by Münkemüller et
al.~(18), which suggested that distributions were positively skewed and
that Blomberg's \(\kappa\) might not behave as a statistic that follows
a normal distribution. However, because their simulations used a
weighted combination of simulated phylogenetic signal strengths, strong
inferences are not possible (and distributional attributes were not the
intended result of their simulations). Thus, for both Pagel's
\(\lambda\) or Blomberg's \(\kappa\), evaluation of statistical moments
across a range of \(\lambda\) used to generate data would be valuable
for adjudicating the --relibalilty-- RELIABILITY of these statistics.
Furthermore, --these are statistics that appear to have expected values
that vary with tree size-- the expected values of these statistics
appear to vary with tree size (18), making comparisons across studies
challenging. Therefore, transformation of these statistics into
\(Z\)-scores would allow evaluation of the efficacy of each statistic to
yield effect sizes that could be used for comparisons of the strength of
phylogenetic signal across traits and lineages.

@@

\textbf{Lambda (\(\lambda\)) estimates of phylogenetic signal are
innacurate.} Computer simulations reveal that for \(\hat{\lambda}\), the
distributional expectations of a Bernoulli variable were mostly upheld.
First, the mean value of \(\hat{\lambda}\) increases as \(\lambda\)
increases, but it is negatively-biased (particularly for small tree
sizes), and is consistently less than the input \(\lambda\) value across
most of its range (Fig. 1 black line). Second, the standard deviation of
\(\hat{\lambda}\) is largest at intermediate values of \(\lambda\) and
smallest at extreme values , implying that the precision in estimating
\(\lambda\) varies across the range of input values (Fig. 1 red line).
Additionally, standard deviations of \(\hat{\lambda}\) are negatively
associated with tree size, and for tress of 128 species or less,
\(\hat{\lambda}\) are quite variable, except for cases when \(\lambda\)
is near or equal to \(1\). Third, the distributions of \(\hat{\lambda}\)
are not normal across its range, but become increasingly skewed at more
extreme values of \(\lambda\) (Fig. 1 blue line). For small tree sizes,
it is also clear that distributions are more platykurtic at intermediate
values of \(\hat{\lambda}\). Taken together these results reveal that
\(\hat\lambda\) inconsistently estimates phylogenetic signal, both
across tree sizes and across the range of input values. Additional
simulations (Supplemental Information) reveal that incorporating
\(\hat\lambda\) in PGLS anova and regression does not adversely affect
the statistical properties of PGLS parameter estimation or model
evaluation (type I error, power, bias in coefficients). Thus, it is
reasonable to incorporate \(\hat\lambda\) in PGLS as a parameter for
tuning the degree of phylogenetic signal in the dependent variables
during the analysis. However, the statistical properties shown in Fig. 1
demonstrate that \(\lambda\) is unsuitable as an effect size for
measuring the strength of phylogenetic signal in data, or for comparing
that signal across datasets.

@@

\textbf{Kappa (\(\kappa\)) estimates of phylogenetic signal are more
reliable.} Simulation results for \(\hat\kappa\) demonstrate that
\(\hat\kappa\) displays better statistical properties. First, as
expected, mean values of \(\hat\kappa\) increase with increasing signal
(\(\lambda\)) irrespective of tree size, though the increase does not
scale linearly with input levels of phylogenetic signal (Fig. 2 black
line). Additionally, the standard deviation of \(\hat\kappa\) is
consistent across tree sizes (Fig. 2 red line), and while it increases
with \(\lambda\), it is always less than the corresponding mean. This
finding is perhaps unsurprising, as \(\hat\kappa\) is lower-bounded by
0, and is never large for small values of \(\lambda\). Importantly,
\(\hat\kappa\) is normally distributed across the range of input
\(\lambda\), and remains consistent in this pattern regardless of tree
size (Fig. 2 blue line). In fact, distributions of \(\hat\kappa\)
display slight skewing only at small tree sizes and when \(\lambda\) is
near or equal to \(1\). This result differs from those of (18), where
skewing appears to be from combining random values generated
independently. These findings reveal that while \(\kappa\) is more
reliable as an estimate of phylogenetic signal, the non-linear scaling
with input signal implies that it should not be considered an effect
size that measures the strength of phylogenetic signal on a common scale
for comparison across datasets.

@@

\textbf{Effect sizes from \(\kappa\) (\(Z_{\kappa}\)) better
characterize phylogenetic signal.} To measure the strength of
phylogenetic signal on a common scale, here we propose effect sizes
(Z-scores) for both \(\lambda\) and \(\kappa\). Statistically, a
standardized effect size may be found as:

@@

Here we evaluate the ability of \(Z_{\lambda}\) and \(Z_{\kappa}\) to
characterize known levels of phylogenetic signal. Both \(Z_{\lambda}\)
and \(Z_{\kappa}\) are associated with input phylogenetic signal
(\(\lambda\)), indicating that both statistics capture the observed
signal (Fig. 3). However, effect sizes from \(\hat{\lambda}\) made
little sense, as they are more strongly associated with tree size than
they are with the actual phylogenetic signal in the data (Fig. 3). By
contrast, \(Z_{\kappa}\) is much more consistent accross tree sizes, and
increases more linearly with incrEasing levels of phylogenetic signal.
Additionally, \(Z_{\kappa}\) exhibits a much stronger association with
phylogenetic signal strength as compared to tree size (Fig. 3), and its
standard deviation across input signal is more consistent. This implies
that similar levels of precision are found with \(Z_{\kappa}\) across
the range of input values. Thus between the two statistics,
\(Z_{\kappa}\) is a more reliable measure of the strength of
phylogenetic signal, and may be used to compare levels of phylogenetic
signal across datasets.

@@

\textbf{Simulations.} Simulations were conducted by generating
pure-birth phylogenies at each of six different tree sizes
(\(n=2^5, 2^6, \cdots, 2^{10}\)), and with differing levels of
phylognetic signal (\(\lambda=0.0, 0.5, \cdots, 1.0\)). We generated 100
random trees for each intersection of tree size and \(\lambda\). For
each \(\lambda\) within each tree size, continuous traits were then
simulated on each phylogeny under a BM model of evolution. For each set
of 100 trees we measured the mean values of \(\hat{\lambda}\) and
\(\kappa\), their standard deviation, and calculated the Shapiro-Wilk
\(W\) statistic as a departure from normality (symmetry). For the
latter, a value of \(1.0\) indicates normally distributed values, while
departures from \(1.0\) indicate skewness. Simulations were then
repeated for both --bananced-- BALANCED and pectinate trees, which
yielded qualitatively similar results (see Supporting Information).
Trees containing polytomies, and an evaluation of \(\hat{\lambda}\) from
models of linear regression and phylogenetic ANOVA, were also
investigated, and results were qualitatively similar to those reported
above (see Supporting Information).

\textbf{Empirical Data.} Surface area to volume ratios (SA:V) and
relative body width (\(\frac{BW}{SVL}\)) measures were obtained --from
as species means-- from individuals of 305 species, from which species
means were obtained (39, 40). A time-dated molecular phylogeny for the
group (45) was pruned to match the species in the phenotypic dataset.
The phylogenetic signal in each trait was then characterized using
\(\kappa\), which was converted to its effect size (\(Z_\kappa\)) using
\texttt{geomorph} 3.3.1 (46, 47), and routines by the authors
(\textbf{to be incorporated in \texttt{geomorph} upon manuscript
acceptance}).

>>>>

The appeal of Pagel's \(\lambda\) and Blomberg's \(\kappa\) as
descriptive statistics is that they provide a basis for interpreting
``weak'' versus ``strong'' phylogenetic signal; i.e., small versus large
values of \(\hat{\lambda}\) or \(\kappa\), respectively, in a
comparative sense (28--30). Nonetheless, an important question that has
yet to be considered is whether such comparisons are analytically
appropriate, and whether these statistics are, or can be, converted to
effect sizes for comparative analyses across datasets. To be statistics
representing phylogenetic signal, they should have reliable
distributional properties, which could be revealed with simulation
experiments. For instance, as a proportional random variable bounded by
\(0\) and \(1\), we might expect that \(\hat{\lambda}\) follows a
Bernoulli distribution (\textbf{add ref}); i.e., branch lengths in a
tree are scaled proportionally to the probability that data arise from a
BM process. Given a known \(\lambda\) value used to generate random data
on a tree, we would also expect that the mean of an empirical sampling
distribution of \(\hat{\lambda}\) would approximately equal \(\lambda\);
the dispersion of \(\hat{\lambda}\) would be largest at intermediate
values of \(\lambda\), \(\hat{\lambda}\) would be predictable over the
range of \(\lambda\) with respect to tree size; the distribution of
\(\hat{\lambda}\) would be symmetric at intermediate values of
\(\lambda\) and more skewed toward values of 0 or 1; and that the
distribution of \(\hat{\lambda}\) will be more platykurtic at
intermediate values of \(\lambda\), becoming more leptokurtic toward 0
and 1 (\textbf{add same ref}). Prior work (18) seems to support some of
these conjectures, based superficially on statistical moments for a
given tree size (mean, variance, skewness, and kurtosis; see Fig. 2 of
ref. (18)). However, because the ``strength of Brownian motion'' was
simulated as a varied weighted-average of data simulated on trees with
\(\lambda=0\) and \(\lambda=1\) and not as prescribed values of
\(\lambda\) (18), interpretation of these patterns is challenging.

By contrast, for Blomberg's \(\kappa\), which is positively unbounded,
we might expect that for any \(\lambda\) used to generate data,
estimates of \(\kappa\) might follow a normal distribution, with values
distributed symmetrically about the input value. This attribute seemed
less reasonable based on the simulations performed by Münkemüller et
al.~(18), which suggested that distributions were positively skewed and
that Blomberg's \(\kappa\) might not behave as a statistic that follows
a normal distribution. However, because their simulations used a
weighted combination of simulated phylogenetic signal strengths, strong
inferences are not possible (and distributional attributes were not the
intended result of their simulations). Thus, for both Pagel's
\(\lambda\) or Blomberg's \(\kappa\), evaluation of statistical moments
across a range of \(\lambda\) used to generate data would be valuable
for adjudicating the reliability of these statistics as effect sizes.
Furthermore, the expected values of these statistics appear to vary with
tree size (18), making comparisons across studies challenging.
Therefore, transformation of these statistics into \(Z\)-scores would
allow evaluation of the efficacy of each statistic to yield effect sizes
that could be used for comparisons of the strength of phylogenetic
signal across traits and lineages.

@@

\textbf{Lambda (\(\lambda\)) estimates of phylogenetic signal are
innacurate.} Computer simulations reveal that for \(\hat{\lambda}\), the
distributional expectations of a Bernoulli variable were mostly upheld.
First, the mean value of \(\hat{\lambda}\) increases as \(\lambda\)
increases, but it is negatively-biased (particularly for small tree
sizes), and is consistently less than the input \(\lambda\) value across
most of its range (Fig. 1 black line). Second, the standard deviation of
\(\hat{\lambda}\) is largest at intermediate values of \(\lambda\) and
smallest at extreme values (Fig.1 red line), implying that the precision
in estimating \(\lambda\) varies across the range of input values.
Additionally, standard deviations of \(\hat{\lambda}\) are negatively
associated with tree size, and for tress of 128 species or less,
\(\hat{\lambda}\) are quite variable, except for cases when \(\lambda\)
is near or equal to \(1\). Third, the distributions of \(\hat{\lambda}\)
are not normal across its range, but become increasingly skewed at more
extreme values of \(\lambda\) (Fig. 1 blue line). For small tree sizes,
it is also clear that distributions are more platykurtic at intermediate
values of \(\hat{\lambda}\). Taken together these results reveal that
\(\hat\lambda\) inconsistently estimates phylogenetic signal, both
across tree sizes and across the range of input values. Additional
simulations (Supplemental Information) reveal that incorporating
\(\hat\lambda\) in PGLS anova and regression does not adversely affect
the statistical properties of PGLS parameter estimation or model
evaluation (type I error, power, bias in coefficients). Thus, it is
reasonable to incorporate \(\hat\lambda\) in PGLS as a parameter for
tuning the degree of phylogenetic signal in the dependent variables
during the analysis. However, the statistical properties shown in Fig. 1
demonstrate that \(\lambda\) is unsuitable as an effect size for
measuring the strength of phylogenetic signal in data, and thus
\(\lambda\) should not be used for comparing phylogenetic signal across
datasets.

@@

\textbf{Kappa (\(\kappa\)) estimates of phylogenetic signal are more
reliable.} Simulation results for \(\hat\kappa\) demonstrate that
\(\hat\kappa\) displays better statistical properties. First, as
expected, mean values of \(\hat\kappa\) increase with increasing signal
(\(\lambda\)) irrespective of tree size, though the increase does not
scale linearly with input levels of phylogenetic signal (Fig. 2 black
line). Additionally, the standard deviation of \(\hat\kappa\) is
consistent across tree sizes (Fig. 2 red line), and while it increases
with \(\lambda\), it is always less than the corresponding mean. This
finding is perhaps unsurprising, as \(\hat\kappa\) is lower-bounded by
0, and is never large for small values of \(\lambda\). Importantly,
\(\hat\kappa\) is normally distributed across the range of input
\(\lambda\), and remains consistent in this pattern regardless of tree
size (Fig. 2 blue line). This result differs from those of (18), where
the skewing appears to be the result of combining random values
generated independently, rather than being a property of \(\kappa\)
itself. Overall, these findings reveal that while \(\kappa\) is more
reliable as an estimate of phylogenetic signal, the non-linear scaling
with input signal implies that it should not be considered an effect
size that measures the strength of phylogenetic signal on a common scale
for comparison across datasets.

@@

\textbf{Effect sizes from \(\kappa\) (\(Z_{\kappa}\)) better
characterize phylogenetic signal.} To measure the strength of
phylogenetic signal on a common scale, we propose effect sizes
(Z-scores) for both \(\lambda\) and \(\kappa\). Statistically, a
standardized effect size may be found as:

@@

Here we evaluate the ability of \(Z_{\lambda}\) and \(Z_{\kappa}\) to
characterize known levels of phylogenetic signal. Both \(Z_{\lambda}\)
and \(Z_{\kappa}\) are associated with input phylogenetic signal
(\(\lambda\)), indicating that both statistics capture the observed
signal (Fig. 3). However, effect sizes from \(\hat{\lambda}\) made
little sense, as they are more strongly associated with tree size than
they are with the actual phylogenetic signal in the data (Fig. 3). By
contrast, \(Z_{\kappa}\) is much more consistent across tree sizes, and
increases more linearly with increasing levels of phylogenetic signal.
Additionally, \(Z_{\kappa}\) exhibits a much stronger association with
phylogenetic signal strength as compared to tree size (Fig. 3), and its
standard deviation across input signal is more consistent. This implies
that similar levels of precision are found with \(Z_{\kappa}\) across
the range of input values. Thus between the two statistics,
\(Z_{\kappa}\) is a more reliable measure of the strength of
phylogenetic signal, and may be used to compare levels of phylogenetic
signal across datasets.

@@

\textbf{Simulations.} Simulations were conducted by generating
pure-birth phylogenies at each of six different tree sizes
(\(n=2^5, 2^6, \cdots, 2^{10}\)), and with differing levels of
phylognetic signal (\(\lambda=0.0, 0.5, \cdots, 1.0\)). We generated 100
random trees for each intersection of tree size and \(\lambda\). For
each \(\lambda\) within each tree size, continuous traits were then
simulated on each phylogeny under a BM model of evolution. For each set
of 100 trees we measured the mean values of \(\hat{\lambda}\) and
\(\kappa\), their standard deviation, and calculated the Shapiro-Wilk
\(W\) statistic as a departure from normality (symmetry). For the
latter, a value of \(1.0\) indicates normally distributed values, while
departures from \(1.0\) indicate skewness. Simulations were then
repeated for both balanced and pectinate trees, which yielded
qualitatively similar results (see Supporting Information). Trees
containing polytomies, and an evaluation of \(\hat{\lambda}\) from
models of linear regression and phylogenetic ANOVA, were also
investigated, and results were qualitatively similar to those reported
above (see Supporting Information).

\textbf{Empirical Data.} Surface area to volume ratios (SA:V) and
relative body width (\(\frac{BW}{SVL}\)) measures were obtained from
individuals of 305 species, from which species means were obtained (39,
40). A time-dated molecular phylogeny for the group (45) was pruned to
match the species in the phenotypic dataset. The phylogenetic signal in
each trait was then characterized using \(\kappa\), which was converted
to its effect size (\(Z_\kappa\)) using \texttt{geomorph} 3.3.1 (46,
47), and routines by the authors (\textbf{to be incorporated in
\texttt{geomorph} upon manuscript acceptance}).
